### 2021-05-05 doc/VerticalArrangements.tex
incorporated Peggys edits in response to EAG clearance (Russ)

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title Page
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff, by the Federal Reserve Board of Governors, or by the Federal Trade Commission.  Furthermore, the views expressed here should not be purported to reflect those of the U.S. Department of Justice.}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Body of Paper
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Given the many different ways a firm may be organized, it is rare that any given merger can be neatly categorized as purely ``horizontal'' or ``vertical.''  Indeed, as the 2020 U.S. Department of Justice (DOJ) and Federal Trade Commission (FTC) Vertical Merger Guidelines (henceforth, the ``VMG'') state, mergers often have both horizontal and vertical aspects.  However, the typical theories and models used to analyze mergers rarely address these complexities.  Horizontal mergers are cast in terms of whether they create significant upward pricing pressure (UPP), while the focus for vertical mergers is on assessing the net effects of the elimination of double marginalization (EDM) versus raising rivals' costs (RRC).  Yet, a given merger may combine all of these effects simultaneously, which raises the necessity of balancing their impacts in a unified framework.

@@

The U.S. waste and recycling industry generates approximately \$80 billion in annual revenues. In recent years the industry has experienced significant merger activity including several large acquisitions between vertically integrated, national competitors. However, solid waste companies tend not to be homogeneous in their degree of vertical integration across geographies. As a result, each of these mergers exhibits a variety of vertical supply chain configurations both pre- and post-merger across their relevant local markets, making it an excellent application for further study of the welfare impacts of mergers that have both horizontal and vertical aspects.

@@

In 2020, Waste Management, the largest waste management company in the U.S., acquired Advanced Disposal Services (ADS), previously the fourth largest company, for \$4.6 billion. GFL Environmental also acquired WCA Waste Corporation for \$1.2 billion in 2020. Republic Services, the second largest waste management company in the U.S., obtained Department of Justice (DOJ) approval to acquire Santek Environmental in March 2021. These three merged companies along with Waste Connections, the third largest waste management company, are estimated to control over 60\% of available landfill capacity nationally and also rank among the top haulers nationwide.\footnote{\tiny{Waste Business Journal, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/ }}} Concentration in local markets varies substantially, however, in both the upstream and downstream markets.

>>>>

\interfootnotelinepenalty=10000

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title Page
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors. Furthermore, the views expressed here should not be purported to reflect those of the Federal Trade Commission or the U.S. Department of Justice.}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Body of Paper
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Given the many different ways a firm may be organized, it is rare that any particular merger can be neatly categorized as purely ``horizontal'' or ``vertical.''  Indeed, as the 2020 U.S. Department of Justice (DOJ) and Federal Trade Commission (FTC) Vertical Merger Guidelines (henceforth, the ``VMG'') state, mergers often have both horizontal and vertical aspects.  However, the typical theories and models used to analyze mergers rarely address these complexities.  Horizontal mergers are cast in terms of whether they create significant upward pricing pressure (UPP), while the focus for vertical mergers is on assessing the net effects of the elimination of double marginalization (EDM) versus raising rivals' costs (RRC).  Yet, a given merger may combine all of these effects simultaneously, which raises the necessity of balancing their impacts in a unified framework.

@@

The U.S. waste and recycling industry generates approximately \$80 billion in annual revenues.\footnote{\tiny{Waste Dive, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/}}} In recent years the industry has experienced significant merger activity including several large acquisitions between vertically integrated, national competitors. However, solid waste companies tend not to be homogeneous in their degree of vertical integration across geographies. As a result, each of these mergers exhibits a variety of vertical supply chain configurations both pre- and post-merger across their relevant local markets, making it an excellent application for further study of the welfare impacts of mergers that have both horizontal and vertical aspects.

@@

In 2020, Waste Management, the largest waste management company in the U.S., acquired Advanced Disposal Services (ADS), previously the fourth largest company, for \$4.6 billion.\footnote{\tiny{Competitive Impact Statement: U.S. and Plantiff States v. Waste Management, Inc. and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1330596/download}}} GFL Environmental also acquired WCA Waste Corporation for \$1.2 billion in 2020.\footnote{\tiny{Waste 360, \url{https://www.waste360.com/business/breaking-gfl-acquires-wca-waste-corp-121-billion}}} Republic Services, the second largest waste management company in the U.S., obtained Department of Justice (DOJ) approval to acquire Santek Environmental in March 2021.\footnote{\tiny{Competitive Impact Statement: U.S. and State of Alabama v. Republic Services, Inc. and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382626/download}}} These three merged companies along with Waste Connections, the third largest waste management company, are estimated to control over 60\% of available landfill capacity nationally and also rank among the top haulers nationwide.\footnote{\tiny{Waste Business Journal, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/ }}} Concentration in local markets varies substantially, however, in both the upstream and downstream markets.
